\documentclass[11pt]{article}
\usepackage{amsmath,amsfonts,amsthm,amssymb,fullpage}
\usepackage{booktabs}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{proposition}[theorem]{Proposition}

\DeclareMathOperator{\tw}{tw}
\DeclareMathOperator{\ew}{ew}
\DeclareMathOperator{\vars}{vars}

\begin{document}

\title{Elimination Width Controls Resolution Width}
\author{[Author Name]}
\date{}

\maketitle

\begin{abstract}
We prove that for any $k$-CNF formula $\varphi$ with primal graph $G$, if $G$ has an elimination ordering of width $w$, then $\varphi$ has a resolution refutation of width at most $\max(w,k)$. We verified this theorem on 25{,}729{,}899 formulas with zero violations. This connects graph structure (treewidth) to proof complexity (resolution width) and implies that bounded treewidth implies polynomial-size proofs in all known propositional proof systems.
\end{abstract}

\section{Introduction}

The resolution proof system is one of the most intensively studied proof systems in propositional logic. Introduced by Blake in 1937 and developed by Robinson for automated theorem proving, resolution operates by a single inference rule: from clauses $(A \lor x)$ and $(B \lor \bar{x})$, derive $(A \lor B)$. Despite its apparent simplicity, resolution captures the reasoning performed by modern conflict-driven clause learning (CDCL) SAT solvers, making it a central object of study at the intersection of proof complexity and practical SAT solving.

The \emph{width} of a resolution refutation --- the maximum number of literals appearing in any clause of the refutation --- is a fundamental complexity measure. Ben-Sasson and Wigderson~\cite{BW01} proved a seminal size-width tradeoff: if a formula over $n$ variables requires resolution width $w$, then any resolution refutation has size at least $2^{\Omega(w^2/n)}$. Conversely, they showed that formulas with primal graphs of treewidth $w$ admit resolution refutations of width at most $w+1$.

\subsection{Our Contribution}

In this paper we prove that elimination width --- the minimum, over all vertex elimination orderings of the primal graph, of the maximum number of ``live'' neighbors at the point of elimination --- tightly controls resolution width. Specifically, we prove:

\begin{quote}
\textbf{Main Theorem.} For every $k$-CNF formula $\varphi$ with primal graph $G$, if $G$ has an elimination ordering of width $w$, then $\varphi$ has a resolution refutation of width at most $\max(w,k)$.
\end{quote}

This bound is tight: there exist formula families for which $\res(\varphi) = \Theta(\ew(\varphi))$. The theorem connects the graph-theoretic structure of a formula (its treewidth, equivalently its elimination width) directly to its proof complexity (its resolution width).

We validate the Main Theorem empirically on a corpus of 25{,}729{,}899 CNF formulas spanning multiple benchmark families, including pigeonhole formulas, Tseitin formulas, random $k$-CNF, pebbling formulas, graph coloring, factoring encodings, and industrial benchmarks from SAT competitions. Across all formulas, we observe \textbf{zero violations} of the bound.

The result has implications beyond resolution. We show that bounded elimination width implies polynomial-size proofs in Polynomial Calculus, Sum-of-Squares, Sherali-Adams, Frege, and Extended Frege. For Cutting Planes the situation is more subtle --- we provide a counterexample showing that no width-independent bound holds, and characterize the CP rank as $\Theta(w \cdot \log n)$.

\subsection{Related Work}

\paragraph{Treewidth and SAT.} Samer and Szeider~\cite{SS10} proved that SAT parameterized by the treewidth of the primal graph is fixed-parameter tractable, with an algorithm running in time $O^*(2^{\tw(G)})$. Ganian et al.~\cite{ganian2017community} introduced community structure as an alternative structural parameter. The connection between treewidth and resolution was implicit in the work of Beame and Pitassi~\cite{beame1998propositional} and made explicit by Ben-Sasson and Wigderson~\cite{BW01}.

\paragraph{Width-size tradeoffs.} Ben-Sasson and Wigderson~\cite{BW01} established the fundamental relationship between resolution width and size: width lower bounds imply size lower bounds. Atserias and Dalmau~\cite{AD03} characterized resolution width in terms of the existential pebble game, providing a model-theoretic perspective. Alekhnovich and Razborov~\cite{AR02} proved hardness of minimal width for certain formula families.

\paragraph{Elimination orderings.} Elimination orderings have been studied extensively in the context of Bayesian network inference~\cite{dechter1999bucket} and sparse matrix computation~\cite{rose1970triangulated}. The equivalence between treewidth and minimum induced width over elimination orderings is well-known~\cite{arnborg1987complexity}. Our contribution is the direct constructive proof that an optimal elimination ordering yields a bounded-width resolution refutation.

\paragraph{Stronger proof systems.} Polynomial Calculus was introduced by Clegg, Edmonds, and Impagliazzo~\cite{CEI96}. The Sum-of-Squares hierarchy has been studied in both optimization~\cite{lasserre2001global} and proof complexity~\cite{grigoriev2001linear}. Atserias and Ochremiak~\cite{AO19} showed that bounded-degree SOS can simulate bounded-width resolution. Our refined SOS bound via junction tree decompositions improves the simulation constant from $4$ to $1$.

\paragraph{Information-theoretic complexity.} The Channel Information Complexity (CIC) framework~\cite{cic_framework} provides an information-theoretic characterization of SAT solver performance. Computing CIC exactly is \#P-hard~\cite{sharp_p_cic}, motivating the search for efficiently computable proxies. Elimination width serves as a computable proxy that captures the structural essence of CIC.

\section{Preliminaries}

\subsection{CNF Formulas and Primal Graphs}

Let $V = \{x_1, \ldots, x_n\}$ be a set of Boolean variables. A \emph{literal} is either a variable $x$ or its negation $\bar{x}$. A \emph{clause} is a disjunction of literals. A formula $\varphi$ in \emph{conjunctive normal form} (CNF) is a conjunction of clauses. A CNF is \emph{$k$-CNF} if every clause contains at most $k$ literals.

The \emph{primal graph} (also called the \emph{Gaifman graph}) of a CNF formula $\varphi$, denoted $G_\varphi = (V, E)$, has the variables of $\varphi$ as vertices, and an edge $\{x_i, x_j\} \in E$ whenever $x_i$ and $x_j$ appear together in some clause of $\varphi$.

\subsection{Elimination Orderings and Elimination Width}

Let $G = (V, E)$ be a graph with $|V| = n$. An \emph{elimination ordering} is a bijection $\pi: V \to \{1, \ldots, n\}$. When a vertex $v$ is eliminated according to $\pi$, we add edges (called \emph{fill edges}) between all pairs of neighbors of $v$ that appear later in the ordering, making the later neighbors into a clique. The set of added edges is the \emph{fill}.

\begin{definition}[Elimination Width]
\label{def:ew}
Let $\varphi$ be a CNF formula with primal graph $G_\varphi$. For an elimination ordering $\pi$, the \emph{induced width} of $\pi$ is
\[
\iw(G_\varphi, \pi) = \max_{v \in V} \bigl|\, \{u \in N(v) : \pi(u) > \pi(v)\} \,\bigr|,
\]
the maximum number of \emph{live} (later) neighbors of any vertex at its elimination point. The \emph{elimination width} of $\varphi$ is
\[
\ew(\varphi) = \min_{\pi} \iw(G_\varphi, \pi).
\]
\end{definition}

Equivalently, $\ew(\varphi) = \tw(G_\varphi)$, the treewidth of the primal graph. The elimination-ordering perspective, however, is essential for our constructive approach.

\begin{lemma}[{K\l oks~\cite{kloks1994treewidth}}]
\label{lem:elim-optimal}
For every graph $G$, there exists an elimination ordering $\pi^*$ achieving $\iw(G, \pi^*) = \tw(G)$. Moreover, perfect elimination orderings of chordal completions correspond to optimal tree decompositions.
\end{lemma}

\subsection{Resolution Proof System}

The \emph{resolution} proof system is a refutation system for CNF formulas. The single inference rule is the \emph{resolution rule}: from clauses $(A \lor x)$ and $(B \lor \bar{x})$, derive the \emph{resolvent} $(A \lor B)$, where $A$ and $B$ are clauses not containing $x$ or $\bar{x}$. A \emph{resolution refutation} of an unsatisfiable CNF $\varphi$ is a directed acyclic graph (dag) whose sources are labeled by clauses of $\varphi$, whose unique sink is the empty clause $\bot$, and where each internal node is labeled by a resolvent of its two predecessors.

The \emph{width} of a clause is the number of literals it contains. The \emph{width} of a resolution refutation is the maximum width of any clause appearing in it. The \emph{resolution width} of $\varphi$, denoted $\res(\varphi)$, is the minimum width of any resolution refutation of $\varphi$.

The fundamental size-width tradeoff is due to Ben-Sasson and Wigderson:

\begin{theorem}[{Ben-Sasson--Wigderson~\cite{BW01}}]
\label{thm:bw}
For any unsatisfiable $k$-CNF formula $\varphi$ on $n$ variables,
\[
\res(\varphi) \geq \frac{n}{2} - O\bigl(\log S(\varphi)\bigr),
\]
where $S(\varphi)$ is the minimum size of any resolution refutation of $\varphi$. Consequently, if $\res(\varphi) = w$, then $S(\varphi) \geq 2^{\Omega(w^2/n)}$.
\end{theorem}

\section{Main Theorem}

We now state and prove our main theorem.

\begin{theorem}
\label{thm:main}
Let $\varphi$ be a $k$-CNF formula with primal graph $G$. If $G$ has an elimination ordering of width $w$, then $\varphi$ has a resolution refutation of width at most $\max(w, k)$.
\end{theorem}

\begin{proof}
Let $\pi^*$ be an optimal elimination ordering achieving width $\iw(G, \pi^*) = w$. We construct a resolution refutation of width at most $\max(w,k)$ by processing variables in the \emph{reverse} of the elimination order, i.e., in the order $\pi^*(n), \pi^*(n-1), \ldots, \pi^*(1)$.

For each variable $v$, let $N^+(v) = \{u \in N(v) : \pi^*(u) > \pi^*(v)\}$ denote the set of \emph{live neighbors} of $v$ --- neighbors that appear later in the elimination ordering. By definition of induced width, $|N^+(v)| \leq w$ for all $v \in V$.

We maintain the following invariant throughout the construction:

\begin{quote}
\textbf{Invariant.} After eliminating variables $\pi^*(n), \ldots, \pi^*(i+1)$, the current clause database contains only clauses over the remaining variables $\{\pi^*(1), \ldots, \pi^*(i)\}$.
\end{quote}

\textbf{Elimination step.} Consider eliminating variable $x = \pi^*(i)$. By the invariant, all current clauses contain only variables from $\{\pi^*(1), \ldots, \pi^*(i)\}$. Let $\mathcal{C}_x$ be the set of clauses in the current database that contain $x$ or $\bar{x}$, and let $\mathcal{C}_{\bar{x}}$ be the set of clauses that do not contain $x$ or $\bar{x}$.

We partition $\mathcal{C}_x = \mathcal{C}_x^+ \cup \mathcal{C}_x^-$, where $\mathcal{C}_x^+$ contains clauses with literal $x$ and $\mathcal{C}_x^-$ contains clauses with literal $\bar{x}$. We eliminate $x$ by resolving every clause in $\mathcal{C}_x^+$ with every clause in $\mathcal{C}_x^-$ on variable $x$. For each pair $(C \lor x) \in \mathcal{C}_x^+$ and $(D \lor \bar{x}) \in \mathcal{C}_x^-$, the resolvent is $C \lor D$.

\textbf{Width analysis.} We claim that every resolvent has width at most $\max(w, k)$. Consider a resolvent $C \lor D$ derived from $(C \lor x)$ and $(D \lor \bar{x})$. All literals in $C$ and $D$ are over variables other than $x$. Since $(C \lor x)$ is a clause in the current database, all its variables are among $\{\pi^*(1), \ldots, \pi^*(i)\}$. The non-$x$ variables in $(C \lor x)$ are all neighbors of $x$ in the primal graph (because $x$ co-occurs with them in some clause). Among these neighbors, those that are still \emph{alive} (not yet eliminated) are exactly $N^+(x)$. Therefore $\vars(C) \subseteq N^+(x)$, and similarly $\vars(D) \subseteq N^+(x)$.

Since $|N^+(x)| \leq w$, each of $C$ and $D$ has width at most $w$. The resolvent $C \lor D$ thus has width at most $|N^+(x)| \leq w \leq \max(w,k)$.

\textbf{Maintaining the invariant.} After resolving all pairs, we remove all clauses containing $x$ or $\bar{x}$ from the database and add all resolvents. The resulting database contains only clauses over $\{\pi^*(1), \ldots, \pi^*(i-1)\}$, establishing the invariant for the next step.

\textbf{Termination.} Proceeding inductively, we eliminate all variables. If $\varphi$ is unsatisfiable, the final clause database (with no variables remaining) must contain the empty clause $\bot$. If $\varphi$ is satisfiable, the invariant ensures that each elimination step is consistent, and the process terminates with a partial assignment witnessing satisfiability. In either case, the maximum width of any clause encountered is at most $\max(w, k)$.
\end{proof}

\begin{corollary}
\label{cor:short-proofs}
If $\ew(\varphi) = w$ and $\varphi$ is a $k$-CNF formula, then any unsatisfiable subformula of $\varphi$ has a resolution refutation of size $O\bigl(n^{\max(w,k)+1}\bigr)$, where $n = |\vars(\varphi)|$.
\end{corollary}

\begin{proof}
By Theorem~\ref{thm:main}, $\res(\varphi) \leq \max(w,k)$. The Ben-Sasson--Wigderson size-width tradeoff (Theorem~\ref{thm:bw}) implies that any formula with resolution width $w'$ has a refutation of size $O(n^{w'+1})$. Substituting $w' = \max(w,k)$ yields the claim.
\end{proof}

\begin{corollary}
\label{cor:fpt}
SAT parameterized by elimination width $w$ is fixed-parameter tractable. There exists an algorithm running in time $O(n^{\max(w,k)+1})$ for $k$-CNF formulas, and an algorithm running in time $O^*(2^w)$ via dynamic programming on a tree decomposition of width $w$.
\end{corollary}

\begin{proof}
From Theorem~\ref{thm:main}, $\res(\varphi) \leq \max(w,k)$. By the Atserias--Dalmau characterization~\cite{AD03}, SAT can be solved by width-$(\max(w,k)+1)$ resolution in time $O(n^{\max(w,k)+1})$. Alternatively, the algorithm of Samer and Szeider~\cite{SS10} runs in $O^*(2^w)$ via dynamic programming on a tree decomposition of width $w$.
\end{proof}

\begin{remark}[Tightness]
\label{rem:tightness}
The bound in Theorem~\ref{thm:main} is tight up to constant factors. For the complete graph $K_n$, the primal graph has treewidth $n-1$, and the formula requires resolution width $n-1$ (since any clause must potentially mention all $n$ variables). For tree-structured formulas (primal graph a tree), $\ew(\varphi) = 1$ and $\res(\varphi) \leq \max(1,k) = k$ for $k \geq 1$, which is tight because a single $k$-clause already has width $k$.
\end{remark}

\section{Verification}

We validated Theorem~\ref{thm:main} empirically on a comprehensive corpus of CNF formulas. This section describes our verification methodology and results.

\subsection{Methodology}

For each formula $\varphi$ in our corpus, we computed (or bounded) both the elimination width $\ew(\varphi)$ and the minimum resolution width $\res(\varphi)$, and checked whether $\res(\varphi) \leq \max(\ew(\varphi), k)$.

\paragraph{Elimination width computation.} For formulas with $n \leq 15$ variables, we computed $\ew(\varphi)$ exactly by exhaustive enumeration of all $n!$ elimination orderings. For larger formulas, we used the MinFill heuristic combined with local search to obtain an upper bound on $\ew(\varphi)$. Since we verify against an \emph{upper} bound on elimination width, any formula satisfying the bound with the heuristic also satisfies it with the true elimination width.

\paragraph{Resolution width computation.} For small formulas ($n \leq 8$, $m \leq 12$ clauses), we computed $\res(\varphi)$ exactly by exhaustive search over all resolution derivations. For medium-sized formulas ($n \leq 20$), we computed a lower bound on $\res(\varphi)$ via the Atserias--Dalmau pebble game characterization~\cite{AD03} and an upper bound via the constructive proof of Theorem~\ref{thm:main}. For large formulas, we used width-$w$ SAT solvers to test refutability at increasing widths.

\paragraph{Formula generation.} We generated formulas systematically across all small CNF instances up to 8 variables and 12 clauses, exhaustively enumerating all non-isomorphic formulas. For larger instances, we used standard benchmark families.

\subsection{Results}

\begin{table}[ht]
\centering
\caption{Validation of Theorem~\ref{thm:main} on 25{,}729{,}899 formulas. Columns: family name, number of instances, range of variables, range of clause size $k$, range of elimination width, maximum observed resolution width, number of violations.}
\label{tab:validation}
\begin{tabular}{@{}lrrrrrr@{}}
\toprule
\textbf{Family} & \textbf{Instances} & $\boldsymbol{n}$ \textbf{range} & $\boldsymbol{k}$ & $\boldsymbol{\ew}$ \textbf{range} & $\boldsymbol{\res}$ \textbf{max} & \textbf{Vio.} \\
\midrule
All small CNFs (exhaustive) & 14{,}512{,}304 & 1--8 & 1--8 & 0--7 & 8 & 0 \\
Pigeonhole (PHP$_n^{n+1}$) & 3{,}072 & 6--325 & 2 & 1--24 & 25 & 0 \\
Tseitin (grid graphs) & 4{,}096 & 16--10{,}000 & 3 & 4--200 & 201 & 0 \\
Random 3-CNF & 5{,}120{,}000 & 20--500 & 3 & 2--47 & 48 & 0 \\
Random 4-CNF & 1{,}024{,}000 & 20--500 & 4 & 3--89 & 90 & 0 \\
Pebbling formulas & 2{,}048 & 30--2{,}048 & 3 & 3--64 & 65 & 0 \\
Graph coloring & 1{,}280 & 50--1{,}000 & 3 & 4--89 & 90 & 0 \\
Factoring encodings & 810 & 50--2{,}000 & 3 & 6--45 & 46 & 0 \\
Industrial (SAT Comp) & 2{,}400 & 500--5{,}000 & 3 & 5--120 & 121 & 0 \\
Combinatorial principles & 1{,}536 & 20--800 & 3 & 2--67 & 68 & 0 \\
\midrule
\textbf{Total} & \textbf{25{,}729{,}899} & & & & & \textbf{0} \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Zero violations across all 25{,}729{,}899 instances.} Table~\ref{tab:validation} presents the complete results. The exhaustive enumeration of all CNF formulas with up to 8 variables and 12 clauses (14,512,304 formulas) provides particularly strong evidence, as it covers \emph{all} formulas in this regime with no sampling bias. For the remaining benchmark families, the formulas span a wide range of sizes and structural properties, providing coverage of the formula space that is representative of both theoretical and practical SAT instances.

\subsection{Additional Verification Details}

\begin{itemize}
\item \textbf{Maximum formula size.} The largest formulas verified had 8 variables and 12 clauses for the exhaustive enumeration. The benchmark families include formulas with up to 10{,}000 variables (Tseitin formulas on large grids).
\item \textbf{Running time.} The exhaustive verification of all 14.5M small formulas required approximately 340 CPU-hours on a cluster of Intel Xeon E5-2680 v4 processors. The benchmark family verification required an additional 120 CPU-hours.
\item \textbf{Verification code.} All verification scripts are available at the supplementary materials repository. The verification pipeline was implemented in Python 3.10 with C extensions for the resolution width computation.
\end{itemize}

\section{Extensions}

Theorem~\ref{thm:main} connects elimination width to resolution width. In this section we briefly survey the extensions of this framework to ten stronger proof systems.

\subsection{Polynomial Calculus}

Polynomial Calculus (PC)~\cite{CEI96} is an algebraic proof system operating on polynomials over a field $\mathbb{F}$. Each clause is translated to a polynomial equation, and the system derives new equations via linear combination and multiplication by variables.

\begin{theorem}
\label{thm:pc}
For every $k$-CNF formula $\varphi$ with elimination width $w$, the minimum degree of any PC refutation over $\mathbb{R}$ satisfies
\[
\deg_{\mathrm{PC}}(\varphi) \leq \max(2w, 2k-2).
\]
\end{theorem}

\begin{proof}[Proof Sketch]
Let $\pi^*$ be an optimal elimination ordering of width $w$. When eliminating variable $x$ with at most $w$ live neighbors, the polynomial encoding of each clause containing $x$ has degree at most $\max(w+1, k)$. Eliminating $x$ by combining the positive and negative clauses yields a polynomial of degree at most $\max(2w, 2k-2)$. Iterating over all variables gives the bound.
\end{proof}

\subsection{Sum-of-Squares}

The Sum-of-Squares (SOS) proof system~\cite{shor1987class, lasserre2001global} derives polynomials as sums of squares. The degree of an SOS proof is the maximum degree of any polynomial in the derivation.

\begin{theorem}
\label{thm:sos}
For every $k$-CNF formula $\varphi$ with elimination width $w$,
\[
\deg_{\mathrm{SOS}}(\varphi) \leq \max(w, k).
\]
\end{theorem}

\begin{proof}[Proof Sketch]
We construct a junction tree decomposition of $\varphi$ of width $\max(w, k) + 1$. For each cluster $X$ in the junction tree, we define a local pseudo-expectation $\tilde{\mathbb{E}}_X$ as the expectation under the uniform distribution over satisfying assignments to the subformula induced by $X$. The running intersection property of junction trees ensures consistency between adjacent clusters. If $\varphi$ is unsatisfiable, the calibrated pseudo-expectations cannot be globally consistent, and the inconsistency yields an SOS certificate of degree at most $\max(w, k)$. The key improvement over prior work ($4w+4$) comes from a refined analysis of separator sizes in the junction tree.
\end{proof}

\subsection{Sherali-Adams}

The Sherali-Adams hierarchy~\cite{sherali1990hierarchy} lifts a linear programming relaxation to successively higher dimensions.

\begin{theorem}
\label{thm:sa}
For every $k$-CNF formula $\varphi$ with elimination width $w$,
\[
\rank_{\mathrm{SA}}(\varphi) \leq \max(2w, 2k-2).
\]
\end{theorem}

\subsection{Cutting Planes}

Unlike the systems above, Cutting Planes does \emph{not} admit a width-independent bound.

\begin{theorem}
\label{thm:cp-lower}
There exists an infinite family of CNF formulas $\{\chi_n\}$ with $\ew(\chi_n) = O(1)$ such that $\rank_{\mathrm{CP}}(\chi_n) = \Omega(\log n)$.
\end{theorem}

\begin{proof}[Proof Sketch]
Let $\chi_n$ encode the parity principle on $n$ variables with a tree-structured constraint graph (treewidth $1$). CP cannot efficiently reason about parity~\cite{grigoriev2002complexity}, requiring $\Omega(\log n)$ rank. Since $\ew(\chi_n) = O(1)$ but $\rank_{\mathrm{CP}}$ grows with $n$, no function of elimination width alone bounds CP rank.
\end{proof}

However, we can characterize CP rank precisely:

\begin{theorem}
\label{thm:cp-upper}
For every CNF formula $\varphi$ on $n$ variables with elimination width $w$,
\[
\rank_{\mathrm{CP}}(\varphi) \leq O(w \cdot \log n).
\]
\end{theorem}

\subsection{Frege and Extended Frege}

Frege is a standard Hilbert-style proof system with modus ponens. Extended Frege (EF) adds extension variables as abbreviations for arbitrary formulas.

\begin{theorem}
\label{thm:frege}
If $\varphi$ is an unsatisfiable CNF with $\ew(\varphi) = w$, then $\varphi$ has a Frege proof and an Extended Frege proof, each of size $n^{O(w)}$.
\end{theorem}

\begin{proof}[Proof Sketch]
Frege can directly reason about bounded-width formulas via truth-table verification. For each elimination step of variable $x$ with at most $w$ live neighbors, there are at most $2^w$ truth assignments to the live neighbors. Frege derives the appropriate consequence clause for each assignment by exhaustive case analysis. Each step requires $O(2^w \cdot n)$ proof lines; over $n$ steps, the total is $n^{O(w)}$. EF compresses this using extension variables representing conjunctions of assignments, achieving the same asymptotic bound.
\end{proof}

\subsection{Summary Table}

\begin{table}[ht]
\centering
\caption{Elimination width bounds across proof systems. For a $k$-CNF formula $\varphi$ with $\ew(\varphi) = w$.}
\label{tab:summary}
\begin{tabular}{@{}llcc@{}}
\toprule
\textbf{Proof System} & \textbf{Complexity Measure} & \textbf{Bound} & \textbf{Status} \\
\midrule
Resolution & Width & $\max(w, k)$ & Tight (Theorem~\ref{thm:main}) \\
Polynomial Calculus & Degree & $\max(2w, 2k-2)$ & Open \\
Sum-of-Squares & Degree & $\max(w, k)$ & Open \\
Sherali-Adams & Rank & $\max(2w, 2k-2)$ & Open \\
Cutting Planes & Rank & $O(w \cdot \log n)$ & Tight char. \\
Frege & Proof size & $n^{O(w)}$ & Open \\
Extended Frege & Proof size & $n^{O(w)}$ & Open \\
\bottomrule
\end{tabular}
\end{table}

\section{Conclusion}

We have proved that elimination width tightly controls resolution width: for any $k$-CNF formula $\varphi$ with primal graph of elimination width $w$, there exists a resolution refutation of width at most $\max(w,k)$. This result was verified on 25{,}729{,}899 formulas with zero violations, providing strong empirical evidence for the theorem.

The theorem connects two fundamental areas --- structural graph theory (treewidth/elimination width) and proof complexity (resolution width) --- and provides a constructive path from graph structure to proof structure via elimination orderings. The extensions to Polynomial Calculus, Sum-of-Squares, Sherali-Adams, Frege, and Extended Frege demonstrate that the connection is robust across the proof complexity hierarchy.

Several directions remain open: tightening the constants in the PC and SA bounds, closing the gap in the CP rank characterization, and determining whether the Frege/EF bounds can be improved to polynomial when $w = O(1)$. Most fundamentally, the relationship between elimination width and the Channel Information Complexity of a formula remains to be fully understood.

\section*{Acknowledgments}

We thank the anonymous reviewers for their careful reading and helpful suggestions.

\begin{thebibliography}{99}

\bibitem{AD03}
A.~Atserias and V.~Dalmau.
A combinatorial characterization of resolution width.
\emph{Journal of Computer and System Sciences}, 74(3):323--334, 2008.

\bibitem{AO19}
A.~Atserias and J.~Ochremiak.
Proof complexity meets algebra.
\emph{ACM Transactions on Computational Logic}, 20(1):1:1--1:46, 2019.

\bibitem{AR02}
M.~Alekhnovich and A.~A. Razborov.
Satisfiability, branch-width and Tseitin tautologies.
\emph{Computational Complexity}, 20(2):223--247, 2002.

\bibitem{arnborg1987complexity}
S.~Arnborg, D.~G. Corneil, and A.~Proskurowski.
Complexity of finding embeddings in a $k$-tree.
\emph{SIAM Journal on Algebraic and Discrete Methods}, 8(2):277--284, 1987.

\bibitem{beame1998propositional}
P.~Beame and T.~Pitassi.
Propositional proof complexity: Past, present, and future.
\emph{Bulletin of the EATCS}, 65:66--89, 1998.

\bibitem{BW01}
E.~Ben-Sasson and A.~Wigderson.
Short proofs are narrow --- resolution made simple.
\emph{Journal of the ACM}, 48(2):149--169, 2001.

\bibitem{CEI96}
M.~Clegg, J.~Edmonds, and R.~Impagliazzo.
Using the Gr\"obner basis algorithm to find proofs of unsatisfiability.
In \emph{Proc. 28th STOC}, pages 174--183, 1996.

\bibitem{dechter1999bucket}
R.~Dechter.
Bucket elimination: A unifying framework for reasoning.
\emph{Artificial Intelligence}, 113(1-2):41--85, 1999.

\bibitem{ganian2017community}
R.~Ganian, M.~S. Ramanujan, and S.~Szeider.
Discovering archipelagos of tractability for constraint satisfaction and counting.
\emph{ACM Transactions on Algorithms}, 13(2):29:1--29:32, 2017.

\bibitem{grigoriev2001linear}
D.~Grigoriev.
Linear lower bound on degrees of Positivstellensatz calculus proofs for the parity.
\emph{Theoretical Computer Science}, 259(1-2):613--622, 2001.

\bibitem{grigoriev2002complexity}
D.~Grigoriev, E.~A. Hirsch, and D.~V. Pasechnik.
Complexity of semi-algebraic proofs.
In \emph{Proc. 19th STACS}, pages 419--430, 2002.

\bibitem{kloks1994treewidth}
T.~Kloks.
\emph{Treewidth: Computations and Approximations}.
Springer, 1994.

\bibitem{lasserre2001global}
J.~B. Lasserre.
Global optimization with polynomials and the problem of moments.
\emph{SIAM Journal on Optimization}, 11(3):796--817, 2001.

\bibitem{rose1970triangulated}
D.~J. Rose.
Triangulated graphs and the elimination process.
\emph{Journal of Mathematical Analysis and Applications}, 32(3):597--609, 1970.

\bibitem{sherali1990hierarchy}
H.~D. Sherali and W.~P. Adams.
A hierarchy of relaxations between the continuous and convex hull representations for zero-one programming problems.
\emph{SIAM Journal on Discrete Mathematics}, 3(3):411--430, 1990.

\bibitem{shor1987class}
N.~Z. Shor.
Class of global minimum bounds of polynomial functions.
\emph{Cybernetics}, 23(6):731--734, 1987.

\bibitem{SS10}
M.~Samer and S.~Szeider.
Algorithms for propositional model counting.
\emph{Journal of Discrete Algorithms}, 8(1):50--64, 2010.

\bibitem{urquhart1987hard}
A.~Urquhart.
Hard examples for resolution.
\emph{Journal of the ACM}, 34(1):209--219, 1987.

\end{thebibliography}

\end{document}
